---
title: "Bayesian workflow with nimbleNeonorm: a simulated example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian workflow: a simulated example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# The MCMC chunks compile C++ code, so they only run where NOT_CRAN is set
# (e.g. devtools::check() and devtools::build_vignettes()).
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  eval = identical(Sys.getenv("NOT_CRAN"), "true")
)
```

This vignette uses simulated data, where the true parameters are known. Real-data case studies are in analysis/

This vignette fits a skewed location-scale model with the MSNBurr and
MSNBurr-IIa distributions, following the steps of the Bayesian workflow of
Gelman et al. (2020): specify the model and its priors, check the priors
through the data they imply, validate the computation on simulated data
with known parameters, check convergence, check the fit against the data,
and compare models.

```{r setup}
library(nimbleNeonorm)
register_neonorm(verbose = FALSE)
```

## 1. The distributions

Both distributions have a location `mu`, a scale `sigma` and a shape
`alpha` (Choir, 2020; Ramadani et al., 2025). Their mode is `mu`, where the
density equals that of a normal distribution with standard deviation
`sigma`. With `alpha = 1` both are symmetric. MSNBurr is skewed to the left
for `alpha < 1` and to the right for `alpha > 1`; MSNBurr-IIa, its mirror
image about `mu`, does the opposite.

```{r densities}
x <- seq(-5, 5, length.out = 401)
dens <- function(f, a) vapply(x, f, numeric(1), mu = 0, sigma = 1, alpha = a)
alphas <- c(0.3, 1, 3)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (fam in c("MSNBurr", "MSNBurr-IIa")) {
  f <- if (fam == "MSNBurr") dmsnburr else dmsnburr2a
  plot(x, dnorm(x), type = "l", lty = 2, ylim = c(0, 0.6),
       xlab = "x", ylab = "density", main = fam)
  for (k in seq_along(alphas)) lines(x, dens(f, alphas[k]), col = k + 1)
  legend("topright", c("normal", paste("alpha =", alphas)),
         lty = c(2, 1, 1, 1), col = c(1, 2:4), bty = "n", cex = 0.8)
}
par(op)
```

## 2. Model and priors

We model `N` observations as independent draws from MSNBurr-IIa. The priors
are weakly informative for data on a unit scale, so the data should be
standardised (or the priors rescaled) before fitting:

* `mu ~ Normal(0, 5)`: the location lies within a few units of zero.
* `sigma ~ Exponential(1)`: a proper prior on the scale that favours
  moderate values without excluding large ones (Gelman, 2006).
* `alpha ~ LogNormal(0, 1)`: its median, 1, is the symmetric case, and it
  treats skewness in either direction alike on the log scale; 95% of its
  mass lies between about 0.14 and 7.1.

```{r model}
model_code <- list(
  normal = nimbleCode({
    for (i in 1:N) y[i] ~ dnorm(mu, sd = sigma)
    mu ~ dnorm(0, sd = 5)
    sigma ~ dexp(1)
  }),
  msnburr = nimbleCode({
    for (i in 1:N) y[i] ~ dmsnburr(mu, sigma, alpha)
    mu ~ dnorm(0, sd = 5)
    sigma ~ dexp(1)
    alpha ~ dlnorm(0, sdlog = 1)
  }),
  msnburr2a = nimbleCode({
    for (i in 1:N) y[i] ~ dmsnburr2a(mu, sigma, alpha)
    mu ~ dnorm(0, sd = 5)
    sigma ~ dexp(1)
    alpha ~ dlnorm(0, sdlog = 1)
  })
)

make_model <- function(family, y) {
  inits <- list(mu = 0, sigma = 1)
  if (family != "normal") inits$alpha <- 1
  nimbleModel(
    model_code[[family]],
    constants = list(N = length(y)),
    data = list(y = y),
    inits = inits
  )
}
```

## 3. Prior predictive check

Before looking at any data, we simulate parameters from the priors and data
from the likelihood, and ask whether the implied data are plausible
(Gelman et al., 2020, Section 2.4; Gabry et al., 2019).

```{r prior-predictive}
n <- 200
prior_model <- make_model("msnburr2a", y = rep(0, n))
prior <- prior_predict_neonorm(prior_model, nsim = 200, seed = 1)

# Spread of each simulated data set
prior_range <- apply(prior$data, 1, function(d) diff(range(d)))
summary(prior_range)

hist(prior$data[1:20, ], breaks = 60, main = "Prior predictive data",
     xlab = "y (20 simulated data sets)")
```

For standardised data, most simulated data sets should span a few to a few
tens of units, with only a small fraction far more extreme. If many
simulated data sets were implausible for the application, the priors would
need to be tightened before fitting; if they were all nearly identical, the
priors would be too restrictive.

## 4. Validating the computation on simulated data

To check that the model and the software recover known parameters, we
simulate data with known values and fit the model to them (Cook et al.,
2006; Gelman et al., 2020, Section 4.1).

```{r fake-data}
true <- c(mu = 0, sigma = 1, alpha = 0.3)   # clearly right-skewed
set.seed(2026)
y <- replicate(n, rmsnburr2a(1, true[["mu"]], true[["sigma"]],
                             true[["alpha"]]))

fit <- runmcmc_neonorm(
  make_model("msnburr2a", y),
  niter = 4000,
  nchains = 4,
  sampler = "slice",
  WAIC = TRUE
)
fit
```

`runmcmc_neonorm()` starts the four chains from different draws of the
prior (`prior_inits_neonorm()`), so that \(\hat{R}\) can detect chains that
have not mixed. The slice sampler is used for `alpha` because it needs no
proposal scale (Neal, 2003).

```{r recovery}
recovery <- cbind(true = true,
                  fit$summary[names(true), c("Mean", "2.5%", "97.5%")])
round(recovery, 3)
```

The true values should lie inside their 95% credible intervals, and the
posterior means should be close to them. A single data set is only a sanity
check, since a 95% interval misses the truth 5% of the time; the
repeated-simulation study in the package repository
(`analysis/simulation-study.R`) estimates bias, mean squared error and
interval coverage over many data sets.

## 5. Convergence diagnostics

The summary reports the rank-normalised split-\(\hat{R}\) and the bulk and
tail effective sample sizes of Vehtari et al. (2021). They recommend
\(\hat{R} < 1.01\) and ESS above 100 per chain; `summarise_neonorm()` warns
otherwise. Trace plots complement these numbers.

```{r convergence}
fit$summary[, c("Rhat", "ESS_bulk", "ESS_tail")]
coda::traceplot(fit$samples[, "alpha"], main = "alpha")
```

## 6. Posterior predictive check

We simulate replicated data sets from the fitted model and compare them
with the observed data, both visually and through a test statistic, the
sample skewness (Gelman et al., 2013, Chapter 6; Gabry et al., 2019).

```{r ppc}
yrep <- posterior_predict_neonorm(fit, ndraws = 200, seed = 1)

plot(density(y), lwd = 2, main = "Observed (black) vs replicated (grey)",
     xlab = "y")
for (i in 1:50) lines(density(yrep[i, ]), col = grDevices::gray(0.6, 0.4))
lines(density(y), lwd = 2)

skewness <- function(v) mean((v - mean(v))^3) / stats::sd(v)^3
t_obs <- skewness(y)
t_rep <- apply(yrep, 1, skewness)
hist(t_rep, breaks = 30, main = "Skewness of replicated data",
     xlab = "sample skewness")
abline(v = t_obs, lwd = 2)
mean(t_rep >= t_obs)   # posterior predictive p-value
```

A posterior predictive p-value far from 0 and 1 means the model reproduces
the skewness of the data.

## 7. Model comparison

Finally, we fit the normal and MSNBurr models to the same data and compare
the three with WAIC (Watanabe, 2010; Vehtari et al., 2017); lower is
better.

```{r compare}
fits <- list(
  normal = runmcmc_neonorm(make_model("normal", y), niter = 4000,
                           WAIC = TRUE),
  msnburr = runmcmc_neonorm(make_model("msnburr", y), niter = 4000,
                            sampler = "slice", WAIC = TRUE),
  msnburr2a = fit
)
waic <- vapply(fits, function(f) f$WAIC$WAIC, numeric(1))
sort(waic)
```

The data were generated from MSNBurr-IIa with a strong right skew, which
MSNBurr-IIa accommodates most easily (Ramadani et al., 2025), so we expect
it to have the lowest WAIC and the normal model, which cannot represent
skewness, the highest. Differences of a few units of WAIC are within its
sampling uncertainty (Vehtari et al., 2017).

```{r cleanup}
cleanup_neonorm(verbose = FALSE)
sessionInfo()
```

## References

Choir, A. S. (2020). *Distribusi neo-normal baru dan karakteristiknya*
[Doctoral dissertation]. Institut Teknologi Sepuluh Nopember.

Cook, S. R., Gelman, A., & Rubin, D. B. (2006). Validation of software for
Bayesian models using posterior quantiles. *Journal of Computational and
Graphical Statistics*, 15(3), 675–692.
https://doi.org/10.1198/106186006X136976

Gabry, J., Simpson, D., Vehtari, A., Betancourt, M., & Gelman, A. (2019).
Visualization in Bayesian workflow. *Journal of the Royal Statistical
Society: Series A*, 182(2), 389–402. https://doi.org/10.1111/rssa.12378

Gelman, A. (2006). Prior distributions for variance parameters in
hierarchical models. *Bayesian Analysis*, 1(3), 515–534.
https://doi.org/10.1214/06-BA117A

Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., &
Rubin, D. B. (2013). *Bayesian Data Analysis* (3rd ed.). CRC Press.

Gelman, A., Vehtari, A., Simpson, D., Margossian, C. C., Carpenter, B.,
Yao, Y., Kennedy, L., Gabry, J., Bürkner, P.-C., & Modrák, M. (2020).
*Bayesian workflow*. arXiv:2011.01808.
https://doi.org/10.48550/arXiv.2011.01808

Neal, R. M. (2003). Slice sampling. *The Annals of Statistics*, 31(3),
705–767. https://doi.org/10.1214/aos/1056562461

Ramadani, E. P., Choir, A. S., Pravitasari, A. A., & Paraguison, J. (2025).
Adding MSNBURR-IIa distribution to MultiBUGS. *Jurnal Aplikasi Statistika &
Komputasi Statistik*, 17(2), 109–130.
https://doi.org/10.34123/jurnalasks.v17i2.804

Vehtari, A., Gelman, A., & Gabry, J. (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC. *Statistics and
Computing*, 27(5), 1413–1432. https://doi.org/10.1007/s11222-016-9696-4

Vehtari, A., Gelman, A., Simpson, D., Carpenter, B., & Bürkner, P.-C.
(2021). Rank-normalization, folding, and localization: An improved
\(\hat{R}\) for assessing convergence of MCMC. *Bayesian Analysis*, 16(2),
667–718. https://doi.org/10.1214/20-BA1221

Watanabe, S. (2010). Asymptotic equivalence of Bayes cross validation and
widely applicable information criterion in singular learning theory.
*Journal of Machine Learning Research*, 11, 3571–3594.
